---
title: "03_Alpha_Div"
Output: html_document
Date created: 04/01/24
Author: Mackenzie Kawahara
Last edited: 04.26.24 #7 on
Overview: Calculations for alpha diversity metrics at rarefied/unrarefied & at Family/ASV levels
---
```{r}
library(Rmisc)
```

## 1. Agglomerate to Family level ##
```{r}
#Based on samples and talking with Alex, ours are likely only able to go down to Family level without losing too much info
#This is only looking at 1000 depth (microbes_rare)
microbes_rare.fam <- tax_glom(microbes_rare, taxrank = "Family", NArm=FALSE)
saveRDS(microbes_rare.fam, "03_Alpha_Div/microbes_rare.fam.rds")
```

## 2. Calculate Shannon, InvSimpson, & Simpson diversity metrics at 1000 depth ##
```{r}
alphaDiv <- estimate_richness(microbes_rare.fam, measures = c("Shannon", "InvSimpson", "Simpson"))

alphaDiv$Samples <-rownames(alphaDiv)

#Combine alpha diversity with sample metadata
microbes_rare.fam@sam_data$Samples <-rownames(microbes_rare.fam@sam_data) #Added 'Samples' column to sam_data to merge with alphaDiv
alphaDiv <- merge(as.matrix(microbes_rare.fam@sam_data), alphaDiv, by = "Samples")

#Save dataframe
readr::write_csv(alphaDiv, here::here("03_Alpha_Div/alphaDiv.csv"))
```

## 3. Plot rough graphs ##
#Shannon Diversity
```{r}
#Plot 1
plot_blank_shan <-  ggplot(alphaDiv, aes(x = as.factor(treatment), y = Shannon, color = treatment)) + 
  scale_color_manual(values = c("#666666", "blue")) + geom_boxplot(lwd = 1.1, outlier.colour = "NA") +
  stat_summary(fun = mean, geom = 'line', mapping = aes(group = treatment, color = treatment), 
               linewidth = 1.25, position = position_dodge(width = 0.9)) #Only box plot, no data points included

#Plot 2
p_cp_nutrients_boxplot <-  ggplot(alphaDiv, aes(x = herbivory, y = Shannon)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y="Shannon Diversity Index")

ggsave(plot=p_cp_nutrients_boxplot, "Graphs/Shannon_cp_nutrients_boxplot.jpeg", units="mm", height = 100, width = 180, scale=1, dpi = 700)

#Plot 3
P1 <- ggplot(alphaDiv,aes(x = as.factor(treatment), y = Shannon, color = treatment)) +
  theme_bw()+
  theme(legend.position = c(.95, .95),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    ) + 
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Shannon Diversity Index")

ggsave(plot=P1, "Graphs/Shannon_P1.jpeg", units="mm", height = 100, width = 180, scale=1, dpi = 700)

#Stats summary using Rmisc package
alpha_summary = summarySE(alphaDiv, measurevar = "Shannon", groupvars = c("herbivory", "treatment"))

#Plot 4 - the average with CI
p_summary = ggplot(alpha_summary, aes(x=herbivory, y=Shannon))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = Shannon - ci, ymax = Shannon + ci), width=0.3)+
  theme_classic()+
  facet_grid(~treatment)+
  labs(x="Consumer Pressure", y="Shannon Diversity Index")

ggsave(plot=p_summary, "Graphs/Shannon_p_summary.jpeg", units="mm", height = 100, width = 180, scale=1, dpi = 700)
```

#Simpson
```{r}
p1_simp <-  ggplot(alphaDiv, aes(x = herbivory, y = Simpson)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y= "Simpson Diversity Index")

p2_simp <- ggplot(alphaDiv,aes(x = as.factor(treatment), y = Simpson, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Simpson Diversity Index")

#Stats summary using Rmisc package
alpha_summary_simp = summarySE(alphaDiv, measurevar = "Simpson", groupvars = c("herbivory", "treatment"))

#Plot with the average with CI
p_summary_simp = ggplot(alpha_summary_simp, aes(x=herbivory, y=Simpson))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = Simpson - ci, ymax = Simpson + ci), width=0.3)+
  theme_classic()+
  facet_grid(~treatment)+
  labs(x="Consumer Pressure", y="Simpson Diversity Index")

ggsave(plot=p_summary_simp, "Graphs/alpha_div_fam/Simpson_p_summary.jpeg", units="mm", height = 100, width = 180, scale=1, dpi = 700)
```

#InvSimpson
```{r}
#Plot 1
p1_invsimp <-  ggplot(alphaDiv, aes(x = herbivory, y = InvSimpson)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y= "Inverse Simpson Diversity Index")

p2_invsimp <- ggplot(alphaDiv,aes(x = as.factor(treatment), y = InvSimpson, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Inverse Simpson Diversity Index")

#Stats summary using Rmisc package
alpha_summary_invsimp = summarySE(alphaDiv, measurevar = "InvSimpson", groupvars = c("herbivory", "treatment"))

#Plot with the average with CI
p_summary_invsimp = ggplot(alpha_summary_invsimp, aes(x=herbivory, y=InvSimpson))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = InvSimpson - ci, ymax = InvSimpson + ci), width=0.3)+
  theme_classic()+
  facet_grid(~treatment)+
  labs(x="Consumer Pressure", y="Inverse Simpson Diversity Index")

ggsave(plot=p_summary_invsimp, "Graphs/alpha_div_fam/InvSimpson_p_summary.jpeg", units="mm", height = 100, width = 180, scale=1, dpi = 700)
```

## 4. Calculate Observed, Chao, & Faith'sPD diversity metric at 5000 depth - add to alphaDiv dataframe ##
```{r}
#Agglomerate to Family level
microbes_rare_5000.fam <- tax_glom(microbes_rare_5000, taxrank = "Family", NArm=FALSE)

alphaDiv_5000 <- estimate_richness(microbes_rare_5000.fam, measures = c("Observed", "Chao1"))
alphaDiv_5000$Samples <-rownames(alphaDiv_5000)

#Combine alpha diversity with sample metadata
microbes_rare_5000.fam@sam_data$Samples <-rownames(microbes_rare_5000.fam@sam_data) #Added 'Samples' column to sam_data to merge with alphaDiv

#alphaDiv <-merge(obs_alphaDiv, alphaDiv, by = "Samples") <-originally did this code, but it didn't include the InvSimpson column so tried the code in line 114. 114 was giving an error, don't know what I did but now when I look at alphaDiv all three diversity metrics are there so I did soemthing right??
alphaDiv <- merge(alphaDiv_5000, alphaDiv, by = "Samples") #This code was giving an error, but now when I look at alphaDiv it has 

#Faith's PD
alphaDiv_FPD <-estimate_richness_wPD(microbes_rare_5000, measures = ("FaithPD"))
alphaDiv_FPD$Samples <-rownames(alphaDiv_FPD)

alphaDiv <-merge(alphaDiv_FPD, alphaDiv, by = "Samples")

#Save dataframe
readr::write_csv(alphaDiv, here::here("03_Alpha_Div/all_alphaDiv.csv"))
```

## 5. Plot rough graphs ##
#Observed Richness
```{r}
p1_obs <-  ggplot(alphaDiv, aes(x = herbivory, y = Observed)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y="Observed Richness")

p2_obs <- ggplot(alphaDiv,aes(x = as.factor(treatment), y = Observed, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Observed Richness")

#Stats summary using Rmisc package
alpha_summary_obs = summarySE(alphaDiv, measurevar = "Observed", groupvars = c("herbivory", "treatment"))

#Plot with the average with CI
p_summary_obs = ggplot(alpha_summary_obs, aes(x=herbivory, y=Observed))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = Observed - ci, ymax = Observed + ci), width=0.3)+
  theme_classic()+
  facet_grid(~treatment)+
  labs(x="Consumer Pressure", y="Observed Richness")
```

#Chao1
```{r}
p1_chao <-  ggplot(alphaDiv, aes(x = herbivory, y = Chao1)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y="Chao1 Richness")

p2_chao <- ggplot(alphaDiv,aes(x = as.factor(treatment), y = Chao1, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Chao1 Richness")

#Stats summary using Rmisc package
alpha_summary_chao = summarySE(alphaDiv, measurevar = "Chao1", groupvars = c("herbivory", "treatment"))

#Plot with the average with CI
p_summary_chao = ggplot(alpha_summary_chao, aes(x=herbivory, y=Chao1))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = Chao1 - ci, ymax = Chao1 + ci), width=0.3)+
  theme_classic()+
  facet_grid(~treatment)+
  labs(x="Consumer Pressure", y="Chao1 Diversity Index")
```

#Faith's PD
```{r}
p1_faith <-  ggplot(alphaDiv, aes(x = herbivory, y = FaithPD)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y="Faith's Phylogenetic Diversity")

p2_faith <- ggplot(alphaDiv,aes(x = as.factor(treatment), y = Chao1, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Faith's Phylogenetic Diversity")

#Stats summary using Rmisc package
alpha_summary_faith = summarySE(alphaDiv, measurevar = "FaithPD", groupvars = c("herbivory", "treatment"))

#Plot with the average with CI
p_summary_faith = ggplot(alpha_summary_faith, aes(x=herbivory, y=FaithPD))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = FaithPD - ci, ymax = FaithPD + ci), width=0.3)+
  theme_classic()+
  facet_grid(~treatment)+
  labs(x="Consumer Pressure", y="Faith's Phylogenetic Diversity Index")
```

## 6. Shannon Diversity - 1000 depth at ASV level ##
```{r}
#AV suggested to look at some measurements unrarefied & at the ASV not fam level
#04.09.24 just chose randomly to start with Shannon

#1000 depth but at ASV Level
alphaDiv_asv <- estimate_richness(microbes_rare, measures = c("Shannon", "InvSimpson", "Simpson"))

alphaDiv_asv$Samples <-rownames(alphaDiv_asv)

#Combine alpha diversity with sample metadata
microbes_rare@sam_data$Samples <-rownames(microbes_rare@sam_data) #Added 'Samples' column to sam_data to merge with alphaDiv
alphaDiv_asv <- merge(as.matrix(microbes_rare@sam_data), alphaDiv_asv, by = "Samples")

#Graph
p1_shan_asv <-  ggplot(alphaDiv_asv, aes(x = herbivory, y = Shannon)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y= "Shannon Diversity Index at the ASV Level")

p2_shan_asv <- ggplot(alphaDiv_asv,aes(x = as.factor(treatment), y = Shannon, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Shannon Diversity Index at the ASV Level")

```

## 7. Shannon Diversity - Unrarefied at family level ##
```{r}
microbes.fam <- tax_glom(microbes.filt, taxrank = "Family", NArm=FALSE)
alphaDiv_unrare <- estimate_richness(microbes.fam, measures = c("Shannon", "InvSimpson", "Simpson"))

alphaDiv_unrare$Samples <-rownames(alphaDiv_unrare)

#Combine alpha diversity with sample metadata
microbes.fam@sam_data$Samples <-rownames(microbes.fam@sam_data) #Added 'Samples' column to sam_data to merge with alphaDiv
alphaDiv_unrare <- merge(as.matrix(microbes.fam@sam_data), alphaDiv_unrare, by = "Samples")

#Graph
p1_shan_unrare_fam <-  ggplot(alphaDiv_unrare, aes(x = herbivory, y = Shannon)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y= "Shannon Diversity Index at the Family Level (Unrarefied)")

p2_shan_unrare_fam <- ggplot(alphaDiv_unrare,aes(x = as.factor(treatment), y = Shannon, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Shannon Diversity Index at the Family Level (Unrarefied)")
```

## 8. Shannon Diversity - Unrarefied at ASV level ##
```{r}
alphaDiv_unrare_asv <- estimate_richness(microbes.filt, measures = c("Shannon", "InvSimpson", "Simpson"))

alphaDiv_unrare_asv$Samples <-rownames(alphaDiv_unrare_asv)

#Combine alpha diversity with sample metadata
microbes.filt@sam_data$Samples <-rownames(microbes.filt@sam_data) #Added 'Samples' column to sam_data to merge with alphaDiv
alphaDiv_unrare_asv <- merge(as.matrix(microbes.filt@sam_data), alphaDiv_unrare_asv, by = "Samples")

#Graph
p1_shan_unrare_asv <-  ggplot(alphaDiv_unrare_asv, aes(x = herbivory, y = Shannon)) + 
  theme_classic()+
  facet_grid(~treatment)+
  geom_boxplot(alpha = 0.6)+
  geom_point(position = "jitter", alpha = 0.5)+
  labs(x="Consumer Pressure", y= "Shannon Diversity Index at the ASV Level (Unrarefied)")

p2_shan_unrare_asv <- ggplot(alphaDiv_unrare_asv,aes(x = as.factor(treatment), y = Shannon, color = treatment)) +
  geom_boxplot(lwd = 1.1, outlier.colour = "NA") + 
  geom_point(position = "jitter", alpha = 0.5) +
  scale_color_manual(values = c("slategray2", "mediumseagreen")) +
  labs(x="Treatment", y="Shannon Diversity Index at the ASV Level (Unrarefied)")
```

